Mirror, mirror on the wall. Which's the best match of them all?

### 2024-01-05 PNAS/index.Rnw
Remove more colored text

\dropcap{I}n forensic \ac{evaluations}, a single conclusion often \ac{relies} on  many comparisons, either implicitly or explicitly.
Multiple comparisons arise persistently when developing statistical methods to address scientific problems \citep{benjaminiControllingFalseDiscovery1995}, and can significantly increase the probability of false "discoveries".  Yet, this potentially significant issue is often ignored.  \ac{Now that vast databases and efficient algorithms are routinely used in forensic evaluations to find "matches" to crime scene items, the problems of close non-matches\citep{pcast} and the erosion of confidence due to multiple comparisons become critically important.}
%Researchers have been focusing on building up databases to develop algorithms for objective evidence comparisons.
%\svp{Then, the algorithms facilitate} searches across (vast) databases, reporting the best $N$ matches to a particular piece of evidence.
%This raises the critically important problem of .

@@

A similar concern was raised in the past with regard to DNA \citep{thompson2003} and latent print evaluations  \citep{koehler2021}: one of the issues \citep{fine2006} leading to the wrongful accusation of Brandon Mayfield in the 2004 Madrid train bombing case was that the large size of the IAFIS database used to search for similar prints made it possible to locate `unusually' close non-matches.
\ac{The probability of finding  close non-matches for an item of evidence is directly proportional to the size of the database available for searching;} false "matches" occur more frequently as databases increase in size and algorithms become more powerful.
% In this paper, we discuss that very close non-matches are far from an unusual phenomenon  but a direct consequence of a large number of comparisons, and simply reflect the power of the search algorithms.

@@

\svp{Compounding this issue,} the use of novel algorithms typically results in a large number of comparisons that may not obvious to the user.
For example, the cross-correlation function\citep{vorburgerApplicationsCrosscorrelationFunctions2011} was one of the first measures proposed for quantifying the similarity  between two patterns in response to the \ac{ 2009 NRC} report \citep{nas2009}, and continues to be used in many pattern searching algorithms to find the best alignment between two images and also to quantify their overall similarity.
Finding the best alignment often consists in sliding one surfaceacross the whole length (for one-dimensional patterns, such as striations) or area (for two dimensional sources, such as impression marks) of the other item while keeping track of the value of a similarity measure.
This mirrors the forensic examination process, where examiners visually rotate or align two surfaces under comparison microscopes.
In order to avoid \ac{problematic} false accusations and a further erosion of the public's trust in science, it is important to address the problem of multiple comparisons by controlling their effect on false discovery rates.

@@

If the blade cuts and the wire edge are scanned at a resolution of $\Sexpr{res*1000} \mu m$, the resulting digital files will require \svp{approximately $\Sexpr{format(round(ncomp_max/1000)*1000, big.mark=',')}$ comparisons per blade cut; $\Sexpr{format(round(ncomp_max/1000)*2000, big.mark=',')}$ comparisons overall}.
These comparisons are required in order to find the optimal alignment between the wire and the blade cut; they are implicit in the calculation of cross-correlation which is the first and often the only step used to objectively assess the similarity between striated evidence such as bullets, aperture shear, and firing pin impressions.
This problem is not unique to algorithms; an examiner would need to physically align the wire and the blade cut by searching along the length of the cut to visually match striations, performing the same process physically that the algorithm performs computationally.
While these sequential comparisons are highly autocorrelated, and we cannot assume sequential independence when calculating the probability of an error, they serve as an upper bound on the number of comparisons which could be performed.

@@

<<control-overall-fpr, fig.cap="Exploration of overall error rate estimates for different values of $r$, the false positive rate for a single comparison. Dotted vertical lines show false positive error rate estimates (from left to right, \\citep{best2022,bajic2020,mattijssen2021}).", fig.width = 8, fig.height = 4, include = F>>=
library(ggplot2)
library(tidyr)
library(dplyr)
library(latex2exp)
library(geomtextpath)
single_rmp = exp(seq(-1000, 0, .1))
N = c(10, 100, 1000, 10000)

@@

Forensic practitioners often report the findings from their examinations in the form of a categorical conclusion reflecting a single decision. This can be misleading when the one decision relies on multiple comparisons which are not individually presented in reports or testimony. In this short contribution, we have shown that the implicit comparisons performed during the forensic analysis of wire cuts have the potential to vastly increase the overall chance that at least one false positive error occurs.
%While often a forensic conclusion is presented as a single decision, underlying that decision are many more comparisons which are not discussed or reported upon.
\ac{Omission of the underlying} details is not necessarily \ac{due to} malice; rather, it results from protocols which focus on finding the best possible match between two items and then evaluating that match, and also from lack of familiarity with basic statistical ideas.

In our example we highlight a simple} case where a wire  is cut using a two-sided blade, but the findings apply to many other situations where a forensic evaluation involves multiple comparisons against a set of accumulated information. Examples include a database search or a situation similar to the example where multiple comparisons of evidence against reference samples are conducted.
%either in a database or in a series of test samples created during the examination process.
As a result, it is critically important that forensic practitioners become aware of the impact of carrying out multiple comparisons on their final conclusions
%science grapple with this multiple comparison problem and
and that they modify lab procedures and testimony accordingly. We suggest three strategies to improve transparency of forensic examinations and to produce more reliable estimates of examination-wide error rates.
%We have three primary suggestions for improvements targeted at making the multiple comparisons visible to everyone involved in the process of evidence evaluation.

>>>>

\dropcap{I}n forensic evaluations, a single conclusion often relies on  many comparisons, either implicitly or explicitly.
Multiple comparisons arise persistently when developing statistical methods to address scientific problems \citep{benjaminiControllingFalseDiscovery1995}, and can significantly increase the probability of false "discoveries".  Yet, this potentially significant issue is often ignored.  Now that vast databases and efficient algorithms are routinely used in forensic evaluations to find "matches" to crime scene items, the problems of close non-matches\citep{pcast} and the erosion of confidence due to multiple comparisons become critically important.
%Researchers have been focusing on building up databases to develop algorithms for objective evidence comparisons.
%\svp{Then, the algorithms facilitate} searches across (vast) databases, reporting the best $N$ matches to a particular piece of evidence.
%This raises the critically important problem of .

@@

A similar concern was raised in the past with regard to DNA \citep{thompson2003} and latent print evaluations  \citep{koehler2021}: one of the issues \citep{fine2006} leading to the wrongful accusation of Brandon Mayfield in the 2004 Madrid train bombing case was that the large size of the IAFIS database used to search for similar prints made it possible to locate `unusually' close non-matches.
The probability of finding  close non-matches for an item of evidence is directly proportional to the size of the database available for searching; false "matches" occur more frequently as databases increase in size and algorithms become more powerful.
% In this paper, we discuss that very close non-matches are far from an unusual phenomenon  but a direct consequence of a large number of comparisons, and simply reflect the power of the search algorithms.

@@

Compounding this issue, the use of novel algorithms typically results in a large number of comparisons that may not obvious to the user.
For example, the cross-correlation function\citep{vorburgerApplicationsCrosscorrelationFunctions2011} was one of the first measures proposed for quantifying the similarity  between two patterns in response to the 2009 NRC report \citep{nas2009}, and continues to be used in many pattern searching algorithms to find the best alignment between two images and also to quantify their overall similarity.
Finding the best alignment often consists in sliding one surfaceacross the whole length (for one-dimensional patterns, such as striations) or area (for two dimensional sources, such as impression marks) of the other item while keeping track of the value of a similarity measure.
This mirrors the forensic examination process, where examiners visually rotate or align two surfaces under comparison microscopes.
In order to avoid problematic false accusations and a further erosion of the public's trust in science, it is important to address the problem of multiple comparisons by controlling their effect on false discovery rates.

@@

If the blade cuts and the wire edge are scanned at a resolution of $\Sexpr{res*1000} \mu m$, the resulting digital files will require approximately $\Sexpr{format(round(ncomp_max/1000)*1000, big.mark=',')}$ comparisons per blade cut; $\Sexpr{format(round(ncomp_max/1000)*2000, big.mark=',')}$ comparisons overall.
These comparisons are required in order to find the optimal alignment between the wire and the blade cut; they are implicit in the calculation of cross-correlation which is the first and often the only step used to objectively assess the similarity between striated evidence such as bullets, aperture shear, and firing pin impressions.
This problem is not unique to algorithms; an examiner would need to physically align the wire and the blade cut by searching along the length of the cut to visually match striations, performing the same process physically that the algorithm performs computationally.
While these sequential comparisons are highly autocorrelated, and we cannot assume sequential independence when calculating the probability of an error, they serve as an upper bound on the number of comparisons which could be performed.

@@

<<control-overall-fpr, fig.cap="Exploration of overall error rate estimates for different values of $r$, the false positive rate for a single comparison. Dotted vertical lines show false positive error rate estimates (from left to right, \\citep{best2022,bajic2020,mattijssen2021}).", fig.width = 8, fig.height = 4, include = F>>=
library(ggplot2)
library(tidyr)
library(dplyr)
# library(latex2exp)
library(geomtextpath)
single_rmp = exp(seq(-1000, 0, .1))
N = c(10, 100, 1000, 10000)

@@

Forensic practitioners often report the findings from their examinations in the form of a categorical conclusion reflecting a single decision. This can be misleading when the one decision relies on multiple comparisons which are not individually presented in reports or testimony. In this short contribution, we have shown that the implicit comparisons performed during the forensic analysis of wire cuts have the potential to vastly increase the overall chance that at least one false positive error occurs.
%While often a forensic conclusion is presented as a single decision, underlying that decision are many more comparisons which are not discussed or reported upon.
Omission of the underlying details is not necessarily due to malice; rather, it results from protocols which focus on finding the best possible match between two items and then evaluating that match, and also from lack of familiarity with basic statistical ideas.

In our example we highlight a simple case where a wire  is cut using a two-sided blade, but the findings apply to many other situations where a forensic evaluation involves multiple comparisons against a set of accumulated information. Examples include a database search or a situation similar to the example where multiple comparisons of evidence against reference samples are conducted.
%either in a database or in a series of test samples created during the examination process.
As a result, it is critically important that forensic practitioners become aware of the impact of carrying out multiple comparisons on their final conclusions
%science grapple with this multiple comparison problem and
and that they modify lab procedures and testimony accordingly. We suggest three strategies to improve transparency of forensic examinations and to produce more reliable estimates of examination-wide error rates.
%We have three primary suggestions for improvements targeted at making the multiple comparisons visible to everyone involved in the process of evidence evaluation.
